---
title: "Final Project: The Economics of Football"
author: "Asger, Laurits, Sejer og Nikolaj"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This project revolves around international football and its economics. Specifically are we focusing on the transfer fees between the players from the season of 1924/1925 up until this season of 2024/2025. We are covering the top-10-transfers every season. With that are we looking into the cost development of said transfers. 

First of all are we going to load tidyverse and other necessary packages. 

```{r tidyverse, echo=TRUE, message=FALSE}
library(tidyverse)
library(ggplot2)
library(readr)
library(dplyr)
library(scales)
```

## Loading our data

Afterwards we need to load our own dataset into R.

```{r data, echo=FALSE}
read_csv("data/Final Project.csv",na="NULL",show_col_types = FALSE)
footballtransfer <- read_csv("data/Final Project.csv", na="NULL",show_col_types = FALSE)

```


## Plotting our data

Now we are going to make a basic graph showing the development of the costs using ggplot. To do this we have elected the help of ChatGPT to tidy up our code, so that we have colourful themes and breaks between the seasons in the X-axis. ChatGPT also recommended converting our coloumn "Average" into a numeric value. 

```{r ggplot,echo=FALSE}
# Load the data
footballtransfer <- read_csv("data/Final Project.csv",show_col_types = FALSE)

# Convert Average to numeric safely, so R can read it better.
footballtransfer <- footballtransfer %>%
  mutate(Average = as.numeric(gsub(",", ".", Average))) %>%
  filter(!is.na(Average), !is.na(Season))

# Get every 5th season label, so the X-axis is more readable. 
season_breaks <- footballtransfer$Season[seq(1, nrow(footballtransfer), by = 5)]

# Create the plot
ggplot(data = footballtransfer, aes(x = Season, y = Average)) +
  geom_line(group = 1, color = "blue") +
  geom_point(color = "red") +
  labs(
    title = "Average over Season",
    x = "Season",
    y = "Average"
  ) +
  scale_x_discrete(breaks = season_breaks) +
  scale_y_continuous(labels = label_comma()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


## Getting the statistics

We wanted to get some basic statistics from the graphs. Like the maximum, overall average fee and so on. It is just to gather a overview of the statistics that we are dealing with. To do that are going to use basic R commands from the Data Carpentry, and tidying it in ChatGPT to visualise it better with the cat() command.  

```{r statistics, echo=FALSE, warning=FALSE}
# Convert relevant columns to numeric if they aren't already
footballstats <- c("nr.1", "nr.2", "nr.3", "nr.4", "nr.5",
                     "nr.6", "nr.7", "nr.8", "nr.9", "nr.10")

footballtransfer[footballstats] <- lapply(footballtransfer[footballstats], function(x) as.numeric(gsub(",", "", x)))

# Combine all numeric values into one long vector
all_values <- unlist(footballtransfer[footballstats])

# Remove NA values (if any)
all_values <- na.omit(all_values)

# Showing the top 10 highest value transfers. 
# Combine values from multiple columns into one vector
all_values <- unlist(footballtransfer[c("nr.1", "nr.2", "nr.3", "nr.4", "nr.5", 
                          "nr.6", "nr.7", "nr.8", "nr.9", "nr.10")])

# Convert to numeric (in case there are characters)
all_values <- as.numeric(all_values)

# Remove NA values
all_values <- na.omit(all_values)

# Get the top 10 highest values
top_10_values <- sort(all_values, decreasing = TRUE)[1:10]

# Calculate and print results
cat("Max value:", max(all_values), "\n")
cat("Min value:", min(all_values), "\n")
cat("Average (mean):", mean(all_values), "\n")
cat("Median:", median(all_values), "\n")
print(top_10_values)

```

The max value is 2.22e+08, which also translates to 222.000.000. Along with that we also wanted to show the top 10 transfer fees across all seasons.

## Analyzing time periods

We are splitting the graph into three different time periods, so that the graphs are easier to analyze and for the reader. The data is split and defined as an early period of football from 1924/1925 until 1955/1956. The next time period have we defined as the middle or rising period of football from 1956/1957 until 1989/1990. The last periode have we defined as the modern period of football from 1990/1991 until today in the 2024/2025 season. We made use of ChatGPT, that had already made the graph from the entire data sheet, and asked ChatGPT to split the graph into three sections based on the graph we had made. 

Disclaimer: We have elected to exclude the 1956/1957 season because it includes a major outlier that does not represent the trends in that elected period as "Early period", or for that matter not even the "Middle period". 

```{r two plots, echo=FALSE}
# Load the data
footballtransfer <- read_csv("data/Final Project.csv",show_col_types = FALSE)

# Convert Average to numeric safely
footballtransfer <- footballtransfer %>%
  mutate(Average = as.numeric(gsub(",", ".", Average))) %>%
  filter(!is.na(Average), !is.na(Season))

# Extract the starting year from the Season string
footballtransfer <- footballtransfer %>%
  mutate(StartYear = as.numeric(substr(Season, 1, 4)))

# Split the data into three time periods
period1 <- footballtransfer %>% filter(StartYear >= 1924 & StartYear <= 1956)
period2 <- footballtransfer %>% filter(StartYear >= 1957 & StartYear <= 1989)
period3 <- footballtransfer %>% filter(StartYear >= 1990)

# We omit the 1956/57 season, because the season is an outlier that shadows the rest
period1 <- footballtransfer %>%
  filter(StartYear >= 1924 & StartYear <= 1956) %>%
  filter(!grepl("1956", Season))

# Helper function to create plot
create_plot <- function(data, title) {
  season_breaks <- data$Season[seq(1, nrow(data), by = 2)]
  
  ggplot(data, aes(x = Season, y = Average)) +
    geom_line(group = 1, color = "blue") +
    geom_point(color = "red") +
    labs(
      title = title,
      x = "Season",
      y = "Average"
    ) +
    scale_x_discrete(breaks = season_breaks) +
    scale_y_continuous(labels = label_comma()) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# Create the plots
plot1 <- create_plot(period1, "Early period: Average over Season (1924/25 – 1956/57)")
plot2 <- create_plot(period2, "Middle period: Average over Season (1957/58 – 1989/90)")
plot3 <- create_plot(period3, "Modern period: Average over Season (1990/91 – 2024/25)")

# Print the plots
print(plot1)
print(plot2)
print(plot3)

```

## Playing with percentages

To understand the data better we are going to convert the price changes per year into percentage, so that the data is more comparable. To do that we mutated the data, so that we would have a new coloumn of data with the data percentages. The command "lag(Average)" means the former seasons average, thereafter we calculate the difference and make it a percentile. 



```{r percentages, echo=FALSE, warning=FALSE}
footballtransfer <- footballtransfer %>%
  arrange(StartYear) %>%
  mutate(PercentChange = (Average - lag(Average)) / lag(Average) * 100)

footballtransfer %>%
  select(Season, PercentChange)

```


After loading the percentages per season, we want to visualise it in a graph. The 1924/1925 value is NA, because the data is comparing to the earlier season. We wanted to make a graph that showed the growth and fall in percentile each season. So based on the data calculated above, we had ChatGPT visualise it with a ggplot. Like earlier we wanted it to correspond with the periods that we had set up, so it was split into three. 


```{r percentages_plot, echo=FALSE,warning=FALSE}
# Load and clean data
footballtransfer <- footballtransfer %>%
  mutate(Average = as.numeric(gsub(",", ".", Average))) %>%
  filter(!is.na(Average), !is.na(Season)) %>%
  mutate(
    StartYear = as.numeric(substr(Season, 1, 4)),
    PercentChange = (Average - lag(Average)) / lag(Average) * 100
  ) %>%
  arrange(StartYear)

# Split into 3 periods
period1 <- footballtransfer %>% filter(StartYear >= 1924 & StartYear <= 1956)
period2 <- footballtransfer %>% filter(StartYear >= 1957 & StartYear <= 1989)
period3 <- footballtransfer %>% filter(StartYear >= 1990)

# Helper function to create percentage change plot
plot_percent_change <- function(data, title) {
  season_breaks <- data$Season[seq(1, nrow(data), by = 2)]
  
  ggplot(data, aes(x = Season, y = PercentChange)) +
    geom_col(aes(fill = PercentChange > 0), show.legend = FALSE) +
    scale_fill_manual(values = c("TRUE" = "#2E86AB", "FALSE" = "#E74C3C")) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    labs(
      title = title,
      x = "Season",
      y = "Percent Change (%)"
    ) +
    scale_x_discrete(breaks = season_breaks) +
    theme_minimal(base_size = 13) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# Create the 3 plots
percent1 <- plot_percent_change(period1, "Early period: Percent Change (1924/25 – 1956/57)")
percent2 <- plot_percent_change(period2, "Middle period: Percent Change (1957/58 – 1989/90)")
percent3 <- plot_percent_change(period3, "Modern period: Percent Change (1990/91 – 2024/25)")

# Display the plots
print(percent1)
print(percent2)
print(percent3)

```


## Comparing data from the World GDP and our data. 

We have looked at the World Bank and its data on development of the World's GDP since 1960. We want to compare this data with our own to see if there are any comparable tendencies between the World's GDP and the economics of football. We had made a new dataset, with the help of ChatGPT and the World Bank. Afterwards we used ChatGPT to use the same format on the graph like we had earlier with the transfer fees. 

```{r worldgdp, echo=FALSE,message=FALSE}
worldgdp <- read_csv2("data/World GDP.csv",show_col_types = FALSE)

ggplot(data = worldgdp, aes(x = Year, y = `World GDP (EUR)`)) +
  geom_line(group = 1, color = "blue") +
  geom_point(color = "red") +
  labs(
    title = "World GDP Development",
    x = "Year",
    y = "World GDP (EUR)"
  ) +
  scale_x_continuous(breaks = seq(min(worldgdp$Year), max(worldgdp$Year), by = 3)) +
  scale_y_continuous(labels = label_comma()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Next we wanted to use ChatGPT to overlay the two datasets so it became more easily comparable and gave a nice tidy version of it together.

```{r worldgdpandfootball, echo=FALSE,show_col_types = FALSE}
# Step 1: Merge datasets
combined_data <- inner_join(
  footballtransfer %>% 
    mutate(StartYear = as.numeric(substr(Season, 1, 4))) %>%
    filter(StartYear >= 1960),
  worldgdp,
  by = c("StartYear" = "Year")
)

# Step 2: Rescale World GDP for plotting
# We'll rescale World GDP to be on a similar scale as transfer fees
max_transfer <- max(combined_data$Average, na.rm = TRUE)
max_gdp <- max(combined_data$`World GDP (EUR)`, na.rm = TRUE)

combined_data <- combined_data %>%
  mutate(
    GDP_scaled = `World GDP (EUR)` / max_gdp * max_transfer
  )

# Step 3: Create dual-axis plot
ggplot(combined_data, aes(x = StartYear)) +
  geom_line(aes(y = Average, color = "Average Transfer Fee"), linewidth = 1) +
  geom_line(aes(y = GDP_scaled, color = "World GDP (scaled)"), linewidth = 1, linetype = "dashed") +
  scale_y_continuous(
    name = "Average Transfer Fee (€)",
    labels = label_comma(),
    sec.axis = sec_axis(
      ~ . * (max_gdp / max_transfer),
      name = "World GDP (€)",
      labels = label_comma()
    )
  ) +
  scale_color_manual(
    name = "",
    values = c("Average Transfer Fee" = "red", "World GDP (scaled)" = "blue")
  ) +
  labs(
    title = "Comparison of Football Transfer Fees and World GDP (1960–2024)",
    x = "Year"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )

```


## Financial crisis vs. Football economics

We can see from the graph above that the two economies follow eachother to some extent. So we wanted to explore this futher, and included major econonimic crisis into the graph, so we could getter picture of the influence those had on both the World's GDP and the transfer fees. We have included the Oil Crisis of 1973, the recession in the 1980's, Black Monday in 1987, Asian Financial Crisis in 1997, the Dot-com crash in 2000, the Wall Street crash in 2008, the Eurozone crisis in 2010, COVID-19 in 2020 and the Energy Crisis in 2022. 

We wanted to make two graphs, one similar to the one above and another one where we compare the changes in percentages. So we asked ChatGPT to make the two graphs based on our earlier templates with the percentile changes per season and the one just above. We wanted a X-axis with only the crisis years and a marker in the graph, so that it was tidy and easily readable. 


```{r worldgdpandfootball2, echo=FALSE}
crisis_years <- c(1973, 1980, 1987, 1997, 2000, 2008, 2010, 2020, 2022)

ggplot(combined_data, aes(x = StartYear)) +
  # Main lines
  geom_line(aes(y = Average, color = "Average Transfer Fee"), linewidth = 1) +
  geom_line(aes(y = GDP_scaled, color = "World GDP (scaled)"), linewidth = 1, linetype = "dashed") +

  # Crisis year vertical lines
  geom_vline(
    xintercept = crisis_years,
    linetype = "dashed",
    color = "gray50",
    linewidth = 0.5
  ) +

  # X-axis shows ONLY crisis years
  scale_x_continuous(
    breaks = crisis_years,
    labels = crisis_years
  ) +

  # Y-axis and secondary Y-axis
  scale_y_continuous(
    name = "Average Transfer Fee (€)",
    labels = scales::label_comma(),
    sec.axis = sec_axis(
      ~ . * (max_gdp / max_transfer),
      name = "World GDP (€)",
      labels = scales::label_comma()
    )
  ) +

  # Color and labels
  scale_color_manual(
    name = "",
    values = c("Average Transfer Fee" = "red", "World GDP (scaled)" = "blue")
  ) +
  labs(
    title = "Comparison of Football Transfer Fees and World GDP (1960–2024)",
    x = "Crisis Years"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )


# Next we are making the same graph, but with the percentage changes instead. 
combined_data <- combined_data %>%
  arrange(StartYear) %>%
  mutate(
    TransferChange = (Average - lag(Average)) / lag(Average) * 100,
    GDPChange = (`World GDP (EUR)` - lag(`World GDP (EUR)`)) / lag(`World GDP (EUR)`) * 100
  )

combined_data <- combined_data %>%
  filter(!is.na(StartYear), !is.na(TransferChange), !is.na(GDPChange))

# Plot
ggplot(combined_data, aes(x = StartYear)) +
  geom_line(aes(y = TransferChange, color = "Transfer Fee % Change"), linewidth = 1) +
  geom_line(aes(y = GDPChange, color = "World GDP % Change"), linewidth = 1, linetype = "dashed") +
  
  # Crisis year lines
  geom_vline(
    xintercept = crisis_years,
    linetype = "dotted",
    color = "gray50"
  ) +
  
  # Crisis years as x-axis 
  scale_x_continuous(
    breaks = crisis_years,
    labels = crisis_years
  ) +
  
  # Labels and theme
  labs(
    title = "Crisis Years vs Transfer Fee and World GDP % Changes (1960–2024)",
    x = "Crisis Years",
    y = "Year-over-Year % Change",
    color = ""
  ) +
  scale_color_manual(values = c("Transfer Fee % Change" = "red", "World GDP % Change" = "blue")) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )

```

You can clearly see some tendencies that correlate to eachother. It is especially clear with the graph showcasing the percentile changes, as nearly all economic crisis saw a dip in the transfer fee percent change compared to the year before. The figure that shows the comparison between the fees and World's GDP shows some clear falls in the fees revolving the crisis after the year 2000, but did not look to have an impact beforehand. 
